Containers and R---the Rockerverse and beyond

### 2020-01-17 manuscript.Rmd
shorten AzureContainers

title: "The Rockerverse: Images, Packages, and Applications for Containerization with R"

@@

  - name: Gergely Daroczi
    # https://github.com/daroczig
      - \orcid{0000-0003-3149-8537}
    email: daroczig@rapporter.net

@@

  - name: Charlotte Van Petegem
    # https://github.com/charvp
      - Department WE02
      - Krijgslaan 281, S9
      - 9000 Gent, Belgium
      - \orcid{0000-0003-0779-4897}
    email: charlotte.vanpetegem@ugent.be

@@

The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells. 
R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects. 
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the \texttt{Dockerfile} of \texttt{python3-minimal}.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.
One of the first implementations of having R respond to client reqests were TCP/IP and local socket servers, such as \CRANpkg{Rserve} and \CRANpkg{svSocket}, which allowed other local or remote processes to make binary requests to interact with R without the need to link to R at compile time and connected other programming languages with R.
R could also empower web services since at least 2007, when [rApache](http://www.rapache.net) was presented at the useR! conference \citep{horner_rapache_2013}.
<!-- website seems down, https://web.archive.org/web/20190627101203/http://rapache.net/manual.html -->
rApache runs an R interpreter inside the Apache web server, which makes it possible and very efficient to build scalable R-based APIs and web services.

Such rApache applications have been used at [rapporter.net](http://rapporter.net), transforming parametrized R Markdown-like documents into interactive web applications for reporting.
A similar design was implemented at [CARD.com](https://www.card.com/), where the optimization of the acquisition flow and the real-time analysis of debit card transactions both relied on an efficient combination of Rserve and rApache running in Docker containers:
the R session behind Rserve acted as a read-only, exteremely fast and scalabale, in-memory database for the many concurrent rApache processes responding to the live-scoring requests of various divisions of the company.
Similar dockerized R scripts were responsible for the ETL processes and even the client-facing email, text message and push notification alerts sent in real-time based on card transactions.
The related Docker images were made available at [https://github.com/cardcorp/card-rocker](https://github.com/cardcorp/card-rocker).
The images extend `rocker/r-base`, e.g., with an SSH client and a workaround for being able to mount SSH keys from the host, Pandoc, the Amazon Web Services (AWS) SDK, and Java, which is required by the AWS SDK.
The AWS SDK allowed to run R consumers reading from real-time data processing streams of [AWS Kinesis](https://aws.amazon.com/kinesis/).
The applications were deployed on Amazon Elastic Container Service ([ECS](https://aws.amazon.com/ecs/)).
The main learnings from using R in Docker was the imporance of not only pinning the R package versions via MRAN, but also moving away from Debian testing to an LTS distribution to minimize the risk of breaking the automated builds of the Docker images and production jobs.

A more recent related work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package, which makes it even easier for R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

[RestRserve](http://restrserve.org) is an alternative R web API framework with the main focus on being high-performant and robust.
It also ships automated Docker builds at https://hub.docker.com/r/rexyai/restrserve/dockerfile.

@@

\label{azurecontainers}
**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io).
A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnote{See \emph{"Deploying a prediction service with Plumber"} vignette for details:  \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01\_plumber\_deploy.html}.}.

@@

Another way this idea of a common environment and the partitioning that docker provides is used in teaching is in the **[Dodona](https://dodona.ugent.be)** project.
Dodona is a web platform developed at Ghent University and is used to teach students basic programming skills.
Dodona uses Docker containers to test submissions by students.
This means that both the code testing the students' submissions and the submission itself are executed in a predictable environment, avoiding compatibility issues between the wide variety of configurations used by students.
The containerization is also used to shield the Dodona servers from bad (or even malicious) code: memory, time and I/O limits are used to make sure students can't overload the system.
The web application communicates with containers by sending configuration information as a JSON document over standard input. 
Every Dodona docker image shares a `main.sh` file that passes through this information to the actual testing framework, while setting up some error handling. 
The testing process in the docker containers sends back the test results by writing a JSON document to its standard output channel.
In june 2019 R support was added to Dodona using an image derived from the Rocker `r-base` image that sets up the `runner` user and and `main.sh` file expected by Dodona\footnote{\href{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}{https://github.com/dodona-edu/docker-images/blob/master/dodona-r.dockerfile}}. 
It also installs the packages required for the testing framework and the exercises so that this doesn't have to happen every time a student's submission is evaluated.
The actual testing of R exercises is done using a custom framework loosely based on `testthat`.
During the development of the testing framework it was found that the testthat framework did not provide enough information to its reporter system to send back all the fields required by Dodona to render its feedback.
Right now, multiple statistics courses are developing exercises to automate the feedback for their lab classes.

@@

The ordering of authors following DN as first author is alphabetical. DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DB wrote the section on \pkg{outsider}.
GD contributed the CARD.com use case.
DE wrote the introduction and the section about Containerization and Rocker.
RC \& EH contributed to the section on interfaces for Docker in R.
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
BM wrote the section on research compendia and made the project Binder-ready.
HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
KR wrote the section about \pkg{wholepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
NX wrote the section on \pkg{liftr}.
All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.

>>>>

title: "Containers and R---the Rockerverse and beyond"

@@

@@

@@

The [**Jupyter Docker Stacks**](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells. 
R is also included in the catchall `jupyter/datascience-notebook` image\footnote{\href{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}{https://jupyter-docker-stacks.readthedocs.io/en/latest/using/selecting.html}}.
For example, these images allow users to quickly start a Jupyter Notebook server locally or build their own specialised images on top of stable toolsets.
The R installation is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See `jupyter/datascience-notebook` `Dockerfile` at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile#L47}.}.

@@

**Gigantum** ([http://gigantum.com/](http://gigantum.com/) is a platform for open and decentralized data science with a focus on using automation and user-friendly tools for easy sharing of reproducible computational workflows.
Gigantum builds on the _Gigantum Client_ (running either locally or on a remote server) for development and execution of data-focused _Projects_, which can be stored and shared via the _Gigantum Hub_ or via a zipfile export.
The Client is a user friendly interface to a backend using Docker containers to package, build, and run Gigantum projects. 
It is configured to use a default set of Docker base images ([https://github.com/gigantum/base-images](https://github.com/gigantum/base-images)), and users are able to define and configure their own custom images.
The available images include two with R based on Ubuntu Linux have the [`c2d4u` CRAN PPA](https://launchpad.net/~marutter/+archive/ubuntu/c2d4u3.5/) pre-configured for installation of binary R packages\footnote{\href{https://docs.gigantum.com/docs/using-r}{https://docs.gigantum.com/docs/using-r}}.
The R images vary in the included authoring environment, i.e., Jupyter in `r-tidyverse` or both Jupyter \& RStudio in `rstudio-server`.
The independent image stack can be traced back to the Gigantum environment and its features.
The R images are based on Gigantum's `python3-minimal` image, originally to keep the existing front-end configuration, but also to provide a consistent Python-to-R interoperability.
The `Dockerfile`s also use build args to specify bases, for example for different version of NVIDIA CUDA for GPU processing\footnote{See \href{https://github.com/gigantum/base-images/blob/master/_templates/python3-minimal-template/Dockerfile}{https://github.com/gigantum/base-images/blob/master/\_templates/python3-minimal-template/Dockerfile} for the `Dockerfile` of `python3-minimal`.} so that appropriate GPU drivers can be enabled automatically when supported.
Furthermore, Gigantum's focus lies on environment management via GUI and ensuring a smooth user interaction, e.g., with reliable and easy conflict detection and resolution.
For this reason, project repositories store authoritative package information in a separate file per package, allowing Git to directly detect conflicts and changes.
A Dockerfile is generated from this description that inherits from the specified base image, and additional custom Docker instructions may be appended by users, though Gigantum's default base images do not currently include the `littler` tool, which is used by Rocker to install packages within `Dockerfile`s.
Because of these specifics, instructions from `rocker/r-ubuntu` could _not_ be readily re-used in this image stack (see Section&nbsp;\nameref{conclusion}).
Both approaches enable `apt` as an installation method, and this is exposed via the GUI-based environment management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.
The image build and publication process is scripted with Python and JSON template configuration files, unlike Rocker which relies on plain `Dockerfile`s.
A minor reason in the inception of the images were also project constraints requiring a Rocker-incompatible licensing of the `Dockerfile`s, i.e., the MIT License.

@@

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time.
However, thanks to advancements in R running as a web service, along with with the ability to deploy R in Docker containers, modern enterprises are now capable of having real-time machine learning powered by R.

The backbone of this work is the \pkg{plumber} ([https://www.rplumber.io/](https://www.rplumber.io/)) package.
The plumber package allows R to be used to host web services through RESTful APIs.
When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response.
For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction.
Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

@@

@@

\label{azurecontainers}
**\pkg{AzureContainers}** is an umbrealla package providing interfaces the three services of Microsoft's Azure Cloud, namely [Container Instances](https://azure.microsoft.com/en-us/services/container-instances/) for running individual containers, [Container Registry](https://azure.microsoft.com/en-us/services/container-registry/) for private image distribution, and [Kubernetes Service](https://azure.microsoft.com/en-us/services/kubernetes-service/) for orchestrated deployments.
While a package like \pkg{plumber} provides the infrastructure for turning an R workflow a service, for production purposes it is usually necessary to take into account scalability, reliability and ease of management.
A single container is limited in the volume of requests it can service, and if the container goes down or is occupied with processing a computationally intensive task, the service becomes unavailable.
These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment, e.g., with [Kubernetes](https://kubernetes.io).
A cluster features load-balancing, autoscaling of containers across numerous servers (in the cloud or on premise), and restarting failed ones.
\pkg{AzureContainers} provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure.
On the client side, it provides simple shells to the Docker, and to Kubernetes' `kubectl` and `helm` tools, that are needed for managing Kubernetes clusters.
The package simplifies complex infrastructure management to a number of R funcction calls, given an Azure account with sufficient credit\footnots{See _"Deploying a prediction service with Plumber"_ vignette at \href{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html}{https://cran.r-project.org/web/packages/AzureContainers/vignettes/vig01_plumber_deploy.html} for details.}.

@@

@@

The ordering of authors following DN as first author is alphabetical. DN conceived the article idea, \href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team}, wrote sections not mentioned below, and revised all sections.
DB wrote the section on \pkg{outsider}.
DE wrote the introduction and the section about Containerization and Rocker.
RC \& EH contributed to the section on interfaces for Docker in R.
DC contributed content on Gigantum.
ME contributed to the section on processing and deployment to cloud services.
CF wrote paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and \pkg{r-db}.
BM wrote the section on research compendia and made the project Binder-ready.
HN \& JN co-wrote the section on enterprise production environments (\pkg{googleComputeEngineR}, Google Cloud Run, Kubernetes).
KR wrote the section about \pkg{wholepunch}.
NR wrote paragraphs about shared work environments and GPUs.
LS \& NT wrote the section on Bioconductor.
NX wrote the section on \pkg{liftr}.
All authors approved the final version. This articles was collaboratively written at \href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The \href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}, \href{https://github.com/nuest/rockerverse-paper/commits/master}{commit history}, and \href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues} provide a detailed view on the respective contributions.
